Verification of the CVM algorithm with a Functional Probabilistic Invariant

### 2025-06-18 paper/sec/intro.tex
elan on vars, bold if-then-else

In this paper, we present a new technique for analyzing randomized algorithms which yields a direct and substantially more general proof of the CVM algorithm.
Our approach is very similar to how deterministic algorithms are verified using loop invariants.
The key difference is that our choice of ``loop invariant'' for the randomized streaming algorithm is a functional probabilistic inequality, namely, we consider invariants of the form:
\[
  \expect [ h ] \leq h(c)
\]
where $h$ is allowed to range over a class of functions, and the expectation is taken over the distribution of the state of the algorithm after consuming each stream element.
By first establishing such an invariant for \cref{alg:cvm}, we can then use it (via different choices of $h$) to establish error bounds for the algorithm.
%For the rest of this paper, we explain this technique, its mechanization, and show how it leads to a proof of the CVM algorithm.
We believe the new proof remains accessible at the undergraduate level, albeit with some exposure to mechanized theorem proving.

>>>>

In this paper, we present a new technique for analyzing randomized algorithms which yields a direct and substantially more general proof of the CVM algorithm.
Our approach is very similar to how deterministic algorithms are verified using loop invariants.
The key difference is that our choice of ``loop invariant'' for the randomized streaming algorithm is a functional probabilistic inequality, namely, we consider invariants of the form:
\[
  \expect [ h ] \leq h(c)
\]
where $c$ is a non-negative real constant, $h : \Rnonneg \rightarrow \Rnonneg$ is allowed to range over a class of functions, and the expectation is taken over the distribution of the state of the algorithm after consuming each stream element.
By first establishing such an invariant for \cref{alg:cvm}, we can then use it (via different choices of $h$) to establish error bounds for the algorithm.
%For the rest of this paper, we explain this technique, its mechanization, and show how it leads to a proof of the CVM algorithm.
We believe the new proof remains accessible at the undergraduate level, albeit with some exposure to mechanized theorem proving.
